\documentclass[12pt,a4paper,openright,twoside]{book}
\usepackage{mathtools}
\RequirePackage[utf8]{inputenc}
\RequirePackage{phd-thesis}
\addbibresource[label=biblio]{phd-thesis.bib}

\mainlinespacing{1.241} % line spacing in mainmatter, comment to default (1)

\begin{document}
	
\frontmatter
\input{front}

\begin{abstract}
    %
    \Gls{AI} represents one of humanity's most transformative technological advancements, with roots in the mid‑20th century and rapidly growing societal impact.
    %
    Over the past decade, the field of \gls{NeSy} \gls{AI} -- which seeks to integrate \emph{symbolic} and \emph{sub-symbolic} approaches -- has gained increasing prominence as a promising paradigm for building intelligent systems.
    %
    By combining the structured reasoning capabilities of symbolic \gls{AI} with the adaptability and scalability of sub-symbolic methods, \gls{NeSy} \gls{AI} aims to overcome the inherent limitations of each paradigm, paving the way for more robust and interpretable systems.

    Within this broad and multifaceted landscape, this thesis focuses on two complementary subfields of \gls{NeSy} \gls{AI}: \gls{SKI} and \gls{SKE}.
    %
    These two areas, often seen as \emph{two sides of the same coin}, provide foundational mechanisms for the development of hybrid intelligent systems.
    %
    \Gls{SKI} enables the injection of structured domain knowledge into sub-symbolic models, guiding their behavior, improving generalization, and enhancing trustworthiness.
    %
    \Gls{SKE}, conversely, facilitates the extraction of symbolic, human-understandable information from these models, bridging the gap between opaque internal representations and interpretability.
    %
    Together, \gls{SKI} and \gls{SKE} form a synergistic loop: while \gls{SKI} influences learning by introducing external structure, \gls{SKE} enables inspection, verification, and refinement of that structure, contributing to the creation of systems that are explainable, reliable, and adaptable to real-world complexity.

    This thesis explores the challenges and opportunities of \gls{NeSy} \gls{AI}, with a particular focus on \gls{SKI}, \gls{SKE}, and their role in the engineering of intelligent systems.
    %
    After an overview of the background and state of the art, we identify key challenges in integrating symbolic and sub-symbolic paradigms.
    %
    We then present original contributions in the form of methodologies, algorithms, and tools designed to advance the capabilities of \gls{NeSy} systems.

    The advent of \glspl{LLM} has further transformed the landscape of \gls{AI}, offering unprecedented capabilities for understanding and generating natural language.
    %
    This thesis investigates how \glspl{LLM} can augment \gls{SKI} and \gls{SKE}, providing new avenues for designing systems that learn and adapt autonomously.
    %
    By leveraging these models, we demonstrate how hybrid approaches can address complex challenges -- such as fairness and decision-making in healthcare -- while ensuring interpretability and alignment with ethical principles.

    Finally, we discuss how these advancements align with the broader vision of \gls{NeSy}, while also contributing to the specific goal of this thesis: enabling the development of systems capable of fully autonomous learning.
    %
    These systems integrate the structured reasoning of symbolic \gls{AI}, the adaptability of sub-symbolic models, and the transformative potential of \glspl{LLM}, opening pathways toward more versatile and intelligent applications.
    %
    Such progress not only advances the field but also brings us closer to the distant horizon of general \gls{AI}, where machines can learn, reason, and adapt autonomously.

    \sloppypar
    \textbf{Keywords} -- \emph{\acrlong{AI}, \acrlong{NeSy} \gls{AI}, \acrlong{SKI}, \acrlong{SKE}, \acrlong{LLM}, intelligent systems, autonomous learning.}
\end{abstract}


\begin{dedication}
    %
    ``The work of each individual contributes to a totality, and so becomes an undying part of the totality.
    %
    That totality of human lives -- past and present and to come -- forms a tapestry that has been in existence now for many tens of thousands of years and has been growing more elaborate and, on the whole, more beautiful in all that time.
    %
    % Even the Spacers are an offshoot of the tapestry and they, too, add to the elaborateness and beauty of the pattern.
    %
    [\dots] An individual life is one thread in the tapestry and what is one thread compared to the whole?
    %
    Daneel, keep your mind fixed firmly on the tapestry and do not let the trailing off of a single thread affect you.''

    ---Isaac Asimov, \emph{Robots and Empire}
    %
\end{dedication}

\begin{acknowledgements} % this is optional
    %
    First and foremost, I would like to thank my parents, Barbara and Marcello, for their unconditional love and support throughout my life.
    %
    I thank my beloved wife, Aqila, for the never ending encouragement and for always pushing me to aim at greater goals.
    %
    I am grateful to my entire family for their help during these years.

    I want to thank my supervisor, Andrea, for his wise advice, continuous support, and for giving me the freedom to explore my ideas.
    %
    I thank my close collaborator, and de-facto a co-supervisor, Giovanni, for his invaluable insights, constructive criticism, and for always being available to discuss ideas and challenges.
    %
    I thank all the other researchers I collaborated with during these years, including Gianluca, Sara, Roberta, Andrea A., Andrea R., Federico and many others.
    %
    I also want to mention -- even if I do not explicitly list all the names -- the researchers of the Lab 4.0 for the shared experiences and the fun times we had together.
    %
    Finally, I thank Ana for being my supervisor during my visiting at the University of Oslo, and for her precious guidance and support.

    Life is also outside Academia, actually, life is mostly outside Academia.
    %
    At the very end of this journey I decided to get married, and I want to thank all my friends who made this day unforgettable and help me in the organization.
    %
    I think I almost lost my mind putting together the writing of this thesis with the wedding organization, but I made it, and I am still (probably) sane.
    %
    Aside from the wedding, these years have been full of emotions, trips, cool experiences and tough moments.
    %
    Eliciting everyone would be infeasible, but I want to mention a few of them.
    %
    Alberto, Eugenio, Marco, Enrico, Michele, the researchers at IFI (Oslo university), especially Riccardo and Paul, the students and the people who frequent the Sthop pub.

    This paragraph has been added after the reviewers' comments.
    %
    Research would not be possible without the peer review process, and it is good to acknowledge the effort of those who dedicate their time to review scientific works.
    %
    At the same time it is also good to point out when a review is done just as a formality, without real care for the work being reviewed.
    %
    I particularly thank the first reviewer for its time and for the detailed and constructive feedback, which helped me improve the quality of this thesis.
    %
    Unfortunately, I cannot say the same for the second reviewer, whose comments were very superficial and not much helpful.
    %
    To be honest, it seems a very rushed -- and out of time -- review, that me myself I would not have written even for a workshop paper.

    Last but not least, I want give a special thanks to somebody that usually does not appear in the acknowledgements of a PhD thesis.
    %
    In fact, since the very beginning of my academic career right after the Master's degree, a major event has shaken the world, and it is still going on: the invasion of Ukraine by fascist Russia.
    %
    The free people of Ukraine are fighting every day to defend not only their homeland, but also the values of freedom and democracy that we all cherish, and without which scientific research itself would not be possible as we know it.

    \vspace{1em}
    \begin{flushright}
        \emph{Slava Ukraini!}

        \emph{Heroiam slava!}
    \end{flushright}

\end{acknowledgements}

\sloppy

%----------------------------------------------------------------------------------------
\dominitoc
\tableofcontents   
\listoffigures     % (optional) comment if empty
\listoftables      % (optional) comment if empty
\lstlistoflistings % (optional) comment if empty
\printglossary[type=\acronymtype,nonumberlist]
%\printnoidxglossaries
%----------------------------------------------------------------------------------------

\mainmatter
\glsresetall

\begin{refsection}
\input{chapters/chapter_1_introduction}

%----------------------------------------------------------------------------------------
%--------------------------------------- PART I -----------------------------------------
%----------------------------------------------------------------------------------------

\part{Background}
\label{part:background}

%\begin{refsection}
\input{chapters/chapter_2_artificial_intelligent_systems}
%\input{chapters/chapter_3_artificial_intelligence}
\input{chapters/chapter_3_neuro_symbolic}
%\input{chapters/chapter_5_llm}
%\printbibliography[title=References,heading=bibintoc]
%\end{refsection}

%----------------------------------------------------------------------------------------
%-------------------------------------- PART II -----------------------------------------
%----------------------------------------------------------------------------------------

\part{Methods, Metrics and Platform for SKI \& SKE}
\label{part:engineering-of-ski-ske}

%\begin{refsection}
\input{chapters/chapter_4_ski_methods}
\input{chapters/chapter_5_psyki}
\input{chapters/chapter_6_fairness}
%\printbibliography[title=References,heading=bibintoc]
%\end{refsection}

%----------------------------------------------------------------------------------------
%------------------------------------- PART III -----------------------------------------
%----------------------------------------------------------------------------------------

\part{Engineering of intelligent systems}\label{part:engineering-of-intelligent-systems}

%\begin{refsection}
\input{chapters/chapter_7_nesy_applications}
\input{chapters/chapter_8_autonomous_learning_systems}
%\printbibliography[title=References,heading=bibintoc]
%\end{refsection}

\input{chapters/chapter_9_conclusions}

\printbibliography[title=Reference,heading=bibintoc]
\end{refsection}

\adjustmtc

\end{document}