%! Author = matteomagnini
%! Date = 05/03/25

%----------------------------------------------------------------------------------------
\chapter{Conclusions}
\label{ch:conclusions}
\minitoc
%----------------------------------------------------------------------------------------

This thesis situates itself at the intersection of symbolic and sub-symbolic \gls{AI}, focusing on the integration of both to enhance the capabilities of intelligent systems.
%
The foundation of this work lies in classical \gls{AI} techniques and \gls{KR} such as computational logic and ontologies (\Cref{ch:intelligent-systems}).
%
An \gls{SLR} was conducted to explore the current landscape of \gls{SKI} and \gls{SKE} techniques (\Cref{ch:nesy-ai}) defining a taxonomy to categorize existing approaches and identifying gaps in the literature.
%
A considerable amount of the work carried out in this thesis contributed to the design and development of \gls{SKI} methods, a novel \gls{SKI} framework (\gls{PSyKI}), and evaluation metrics (\Cref{ch:ski-methods-and-contributions,ch:psyki}).
%
Also, studies about \gls{AI} and society, in particular related to the fairness of \gls{AI} systems through \gls{SKI} methods, were conducted (\Cref{ch:fairness-through-ski}).
%
Real-world applications of \gls{SKI} techniques were explored, demonstrating their effectiveness in various domains (\Cref{ch:nesy-ai-for-real-world-applications}).
%
Finally, because the ultimate goal of this work is to design and implement intelligent systems that can autonomously learn knowledge about a given domain, solutions based on \gls{SKI} and \gls{SKE} techniques were proposed and evaluated (\Cref{ch:autonomous-learning-systems}).


\section{Discussion}\label{sec:discussion}

\subsection*{About the relevance of \gls{SKI} and \gls{SKE} techniques}
%
Regarding the research question \Cref{itm:rq0}, introduced in \Cref{ch:introduction}, we conducted an \gls{SLR} to investigate the current state of \gls{SKI} and \gls{SKE} techniques (\Cref{ch:nesy-ai}).
%
The \gls{SLR} investigated \emph{249} relevant papers, identifying \emph{117} distinct \gls{SKI} works and \emph{132} \gls{SKE} methods.
%
These high numbers, along with the increasing trend of publications in recent years, indicate that \gls{SKI} and \gls{SKE} are active research areas within the \gls{AI} community.
%
In the course of other chapters -- e.g., when presenting the possibility to use \gls{SKI} techniques to enhance the fairness of \gls{AI} systems in \Cref{ch:fairness-through-ski} and when showing real-world applications in \Cref{ch:nesy-ai-for-real-world-applications} -- we further demonstrated the relevance of \gls{SKI} and \gls{SKE} techniques in addressing contemporary challenges in \gls{AI}.
%
In \Cref{ch:ski-methods-and-contributions} we also presented several novel \gls{SKI} methods and in \Cref{ch:psyki} a new framework (\gls{PSyKI}) to facilitate the application of \gls{SKI} techniques.


\subsection*{About the characteristics of \gls{SKI} and \gls{SKE} techniques}
%
The \gls{SLR} also provided insights into the characteristics of existing \gls{SKI} and \gls{SKE} techniques posed by the research question \Cref{itm:rq1}.
%
A comprehensive taxonomy was developed to categorize the techniques based on various dimensions.


For \gls{SKI}, the dimensions that characterise most the methods are \emph{input knowledge type}, \emph{sub-symbolic model target}, and \emph{injection strategy}.
%
We observed that \gls{SKI} techniques predominantly utilize structured knowledge representations, such as ontologies and knowledge graphs, to inject knowledge into sub-symbolic models.
%
Also logic rules are widely used.
%
Regarding the target sub-symbolic models, \gls{SKI} methods mainly focus on \glspl{NN} of any kind, with a predominance of feed-forward architectures.
%
Finally, the injection strategies are quite balanced among the three main categories: \emph{structuring}, \emph{embedding}, and \emph{guided learning}.


For \gls{SKE}, the most characterising dimensions are \emph{output knowledge shape}, \emph{output knowledge expressiveness}, and \emph{translucency}.
%
\gls{SKE} techniques predominantly generate knowledge in the form of logic rules, with \gls{DT} being the second most common shape.
%
The expressiveness of the extracted knowledge is often limited to propositional logic, with fewer methods producing first-order logic or more complex representations.
%
Regarding translucency, \gls{SKE} methods are fairly evenly distributed among \emph{pedagogical} and \emph{decompositional} approaches.


\subsection*{Measuring the effectiveness of \gls{SKI} and \gls{SKE} techniques}
%
Accuracy, precision, recall, and F1-score are the most commonly used metrics to evaluate \gls{ML} models, but these metrics do not capture the full spectrum of qualities of \gls{SKI} and \gls{SKE} techniques.
%
To answer the research question \Cref{itm:rq2}, we proposed \gls{QoS} metrics specifically designed to assess other aspects beyond traditional performance metrics (\Cref{ch:ski-methods-and-contributions}).
%
In particular, in \Cref{sec:ski-meets-intelligent-agents} four different \gls{QoS} metrics were proposed: \emph{memory footprint}, \emph{energy consumption}, \emph{latency}, and \emph{data efficiency}.
%
Additionally, in \Cref{sec:empirical-study-on-the-robustness-of-ski-methods}, we introduced a \emph{robustness} metric to evaluate the resilience of \gls{SKI} methods against different types of data degradation.
%
All the metrics were rigorously defined and empirically validated through experiments.
%
Finally, the fairness dimension was explored in \Cref{ch:fairness-through-ski}, where we demonstrated how \gls{SKI} techniques can mitigate the bias of \gls{ML} models, quantifiable through established fairness metrics.


\subsection*{When and where to use \gls{SKI} and \gls{SKE} techniques}
%
In the context of real-world applications, we addressed the research question \Cref{itm:rq3} by exploring various domains and purposes where \gls{SKI} and \gls{SKE} techniques have been successfully applied (\Cref{ch:nesy-ai-for-real-world-applications}).
%
Specifically:
%
\begin{inlinelist}
    %
    \item we mitigated bias in \gls{AI} systems (\Cref{sec:fauci}) using \gls{SKI} techniques to enhance fairness,
    %
    \item we examined applications in healthcare (\Cref{sec:ske-for-explainable-nutritional-recommenders}) using \gls{SKE} to enhance explainability and customization of nutritional recommendations,
    %
    \item we introduced a protocol for multi-agent based explanations in \gls{AI} systems (\Cref{sec:a-general-purpose-protocol-for-multi-agent-based-explanations}) to improve transparency and user trust,
    %
    \item again in the healthcare domain (\Cref{sec:nesy-ai-for-supporting-chronic-disease-diagnosis-and-monitoring}), we used \gls{SKI} to provide established domain knowledge to \gls{ML} predictors in the context of chronic diseases,
    %
    \item finally, we explored the use of \gls{RAG} -- that falls into our \gls{SKI} definition -- by providing contextual prior knowledge to \glspl{LLM}.
    %
\end{inlinelist}


\subsection*{\gls{SKI} and \gls{SKE} techniques for designing \gls{NeSy} \gls{AI} systems}
%
The last research question \Cref{itm:rq4} was addressed in both \Cref{ch:nesy-ai-for-real-world-applications} (already summarised in the previous paragraph) and \Cref{ch:autonomous-learning-systems}.
%
Because the ultimate goal of this thesis is to design and implement intelligent systems that can autonomously learn, in \Cref{ch:autonomous-learning-systems} we focused entirely on this aspect.
%
In \Cref{sec:cycle-ski-ske} we proposed a conceptual architecture for autonomous learning systems that leverage \gls{SKI} and \gls{SKE} techniques to continuously acquire, integrate, and refine knowledge.
%
The idea consists in cycle of \gls{SKI} and \gls{SKE} steps, where knowledge is injected into sub-symbolic models, which are then used to extract new knowledge that can be reintegrated.
%
Concrete systems that automatically learn knowledge are presented in \Cref{sec:llm-as-oracles-for-instantiating-ontologies-with-domain-specific-knowledge,sec:actively-learning-ontologies}.
%
The first work (\Cref{sec:llm-as-oracles-for-instantiating-ontologies-with-domain-specific-knowledge}) uses \glspl{LLM} as oracles to populate ontologies starting from a predefined schema (the schema itself can be asked to the \gls{LLM}).
%
One can perform multiple iterations to refine the ontology at will.
%
The second work (\Cref{sec:actively-learning-ontologies}) instead consists in an active-learning framework that exploits \glspl{LLM} as a teacher and the student learns the target ontology by posing queries to the teacher.
%
Membership queries are directly answered by the \gls{LLM}, while equivalence queries are simulated through the \gls{PAC} framework.
%
These contributions demonstrate the feasibility of designing autonomous learning systems.



\section{Future work}\label{sec:future-work}

\subsection*{Universal \gls{SKI} language}
%
The vast majority of the existing \gls{SKI} methods in the literature suffer from a lack of implementations and standardization.
%
The techniques that actually have an implementation often rely on hard-coded rules and procedures that make them difficult to adapt to different contexts.
%
Moreover, the use of logic programming formalism is often required, which can be a barrier for practitioners not familiar with it.
%
\Gls{PSyKI} (\Cref{sec:psyki}) partially addresses these issues by providing a simpler interface to apply \gls{SKI} techniques, but it still requires the user to know logic programming to define the input knowledge.
%
To address these issues and make \gls{SKI} accessible to a broader audience of users, a simplified and universal \gls{SKI} language could be developed.


\subsection*{Practical implementation of the \gls{SKI}-\gls{SKE} cycle}
%
The conceptual architecture for extracting and refining knowledge through a cycle of \gls{SKI} and \gls{SKE} steps was proposed in \Cref{sec:cycle-ski-ske}.
%
However, a practical implementation of this architecture is still missing in the literature to the best of our knowledge.
%
Developing a concrete system that embodies this cycle, possibly involving multiple intelligent agents sharing knowledge between each others, would be of great interest.


\subsection*{Social implication of \gls{SKI} and \gls{SKE}}
%
Both \gls{SKI} and \gls{SKE} can affect the social implications of \gls{AI} systems.
%
Concerning \gls{SKI}, we demonstrated in \Cref{ch:fairness-through-ski} how these techniques can be used to mitigate bias in \gls{AI} systems.
%
Moreover, they can be used to enhance the transparency and explainability -- by design -- of \gls{AI} models, which are crucial for user trust and acceptance (\Cref{ch:ski-methods-and-contributions}).
%
\Gls{SKE} techniques can also contribute to explainability by providing human-understandable representations of the knowledge learned by sub-symbolic models (\Cref{sec:nesy-ai-for-supporting-chronic-disease-diagnosis-and-monitoring,sec:a-general-purpose-protocol-for-multi-agent-based-explanations}).

Despite having these positive aspects, a measurement of the social impact of \gls{SKI} and \gls{SKE} in real-world applications among the users is still missing.
%
Moreover, \gls{SKI} methods and in particular the most recent techniques that involve the use of \glspl{LLM} raise concerns about data privacy and \gls{AI} safety.
%
Indeed, if the knowledge used to inject into a sub-symbolic model contains sensitive information, and this knowledge is sent to a third party, there is a risk that this information could be inadvertently exposed or misused.
%
Finally, whenever \glspl{LLM} are used, there is always a chance of having hallucinated or biased information that could compromise the integrity of the knowledge being injected.
%
We believe that these aspects should be further investigated in future studies.


\subsection*{Autonomous learning in the real world}
%
Knowledge and data are just a representation of the real world.
%
The best way to learn about the real world is to interact with it, like animals and humans did in millions of years of evolution (billions for the life in general).
%
We believe that the next step for autonomous learning systems -- and ultimately \gls{AGI} -- is to be embodied in the real world, through robots and other physical agents that can perceive and act in the environment.
%
When this will take place, it will be of paramount importance to guarantee human and environmental safety, as well as ethical behavior of the agents.
%
This will require a multidisciplinary approach -- involving computer science, robotics, ethics, lawmakers, and social sciences -- and a strong collaboration between academia, industry, and the society at large.
